---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
```{r include=FALSE, cache=FALSE}
require(ggplot2)
```

## Loading and preprocessing the data
```{r echo = TRUE} 
step_data = read.csv("C:\\Users\\krodriguez\\Documents\\Coursera\\Reproducible Research\\Week 2\\RepData_PeerAssessment1\\activity\\activity.csv", stringsAsFactors = FALSE)
```

## What is mean total number of steps taken per day?
```{r echo = TRUE, comment=NA}
omit_data = na.omit(step_data)

omit_data_sum = aggregate(steps ~ date, omit_data, sum)
omit_data_sum$date = as.Date(omit_data_sum$date)

ggplot(omit_data_sum, aes(x=date, y=steps)) + geom_histogram(stat="identity") + scale_x_date(breaks = waiver()) + ggtitle("Total Number of Steps per Day")
```

####This is the average number of steps by day
```{r echo =TRUE, comment=NA}
print(aggregate(steps ~ date, omit_data, mean))
```

####This is the median number of steps by day
```{r echo =TRUE, comment=NA}
print(aggregate(steps ~ date, omit_data, median))
```


## What is the average daily activity pattern?
```{r echo = TRUE, comment=NA}
omit_data_avg = aggregate(steps ~ interval, omit_data, mean)

ggplot(omit_data_avg, aes(x=interval, y = steps)) + geom_line() + ggtitle("Time Series Plot of the 5 Minute Interval and the Average Number of Steps Taken")
```

####This is the highest 5-minute Interval that on Average Contains the Maximum Number of Steps 
```{r echo =TRUE, comment=NA}
max_interval =  omit_data_avg[which(omit_data_avg$steps == max(omit_data_avg$steps)),]
print(max_interval$interval, row.names = FALSE)
```

## Imputing missing values
#### This is the Total Number of NA values in the data set
```{r echo = TRUE, comment = NA}
sum(is.na(step_data))
```

#### This is the New Dataset with NAs replaced with the Average of Steps for the Whole Data Set
```{r echo = TRUE, comment = NA}
step_data_filled = step_data
step_data_filled$steps[is.na(step_data_filled$steps)] = mean(step_data_filled$steps, na.rm = TRUE)
```
#### This is the Total Number of Steps Taken by Day for the Dataset with NAs replaced 
```{r echo = TRUE, comment = NA}
data_sum= aggregate(steps ~ date, step_data_filled, sum)
data_sum$date = as.Date(data_sum$date)

ggplot(data_sum, aes(x=date, y=steps)) + geom_histogram(stat="identity") + scale_x_date(breaks = waiver()) + ggtitle("Total Number of Steps per Day for New Dataset")
```

#### This is the Mean of Total Number of Steps Taken Each Day for the New Dataset
```{r echo = TRUE, comment = NA}
aggregate(steps ~ date, step_data_filled, mean)
```

#### This is the Median of Total Number of Steps Taken Each Day for the New Dataset
```{r echo = TRUE, comment = NA}
aggregate(steps ~ date, step_data_filled, median)
```
After replacing NA values with the average of steps for the whole dataset, there is a slight effect on the estimates of the total number of steps taken per day.

The most noticeable effect is on the median and average calculations for days with data containing a high number of NAs. More speficially, days 10-1, 10-8, 11-1, 11-4, 11-9, 11-10, 11-14, 11-30 all showed a sharp increase in median from 0 to 37.38. This is because the original dataset contained majority NAs for these specific days. This resulted in the replacement from NA to 37.38 for most entries in those days.


## Are there differences in activity patterns between weekdays and weekends?
```{r echo = TRUE, comment = NA}
step_data_filled$date = as.Date(step_data_filled$date)

days = as.data.frame(weekdays(step_data_filled$date), stringsAsFactors = TRUE )
colnames(days) = "day"

step_data_filled = cbind(step_data_filled, days)

weekend = c("Saturday", "Sunday")
step_data_filled$weekday <- factor((step_data_filled$day %in% weekend), 
         levels=c(TRUE, FALSE), labels=c('weekend', 'weekday')) 

data_avg_weekend = aggregate(steps ~ interval * weekday, step_data_filled, mean)

ggplot(data_avg_weekend, aes(x=interval, y = steps, group = weekday)) + geom_line() + ggtitle("Time Series Plot of the 5 Minute Interval and the Average Number of Steps Taken") + facet_wrap(~weekday , ncol = 1)
```
